Simple Type Theory is not too Simple

### 2021-01-12 article.tex
Add one reference and add comments on topological spaces in conclusion

In the same time, some proof assistants have made impressive progress on the algebra front. Indeed, in 2003 Laurent Chicli formalized sheaves and affine schemes in the proof assistant Coq based on a powerful dependent type theory known as the \emph{Calculus of Inductive Constructions} \cite{chiclithesis}. More recently, in 2019 Grothendieck's schemes have been formalized by the mathematician Kevin Buzzard and three students using the brave new Lean theorem prover also based on a dependent type theory\footnote{\url{https://github.com/ramonfmir/lean-scheme}}. \\

@@

For our formalization we built upon Ballarin's work \cite{ballarin20} formalizing parts of Jacobson's \textit{Basic Algebra} textbook. We believe Ballarin's nice formalization could serve as a first stepping stone to a new algebra library for Isabelle. We also hope the present work could serve as a second stepping stone in that direction. Also, we seized the opportunity of formalizing schemes to partly rebuild the topology library. Our new topology library is entirely built on modern locales instead of using type classes like it is the case in \textit{HOL-Topological\_Spaces.thy}, allowing more abstraction. In particular, our topological spaces have explicit carrier sets instead of using \textit{UNIV}. \\
Our work demonstrates that intricate dependencies of algebraic structures can be managed elegantly in Isabelle. 
Actually, it seems that our experience in Isabelle to formalize schemes was much smoother than the corresponding one in Lean, since Kevin Buzzard reported on his blog the important difficulties he faced during his formalization in Lean:
\begin{quote}
	The project is completely incompatible with modern Lean and mathlib, but if you compile it then you get a sorry-free proof that an affine scheme is a scheme. During our proof we ran into a huge problem because our blueprint, the Stacks Project, assumed that $R[1/f][1/g]=R[1/fg]$, and this turns out to be unprovable for Lean’s version of equality [\dots]. The Lean community wrestled with this idea, and has ultimately come up with a beefed-up notion of equality for which the identity is now true.\footnote{\url{https://xenaproject.wordpress.com/2020/06/05/the-sphere-eversion-project/}}
\end{quote}	
For the first author of the present work, well-versed in dependent type theory, it raises the question of whether dependent types could be, at least in some situations, more of a burden than an advantage. Indeed, in the age of machine learning one may suspect there is a trade-off between the expressiveness of a type theory and one's ability to subject it to automation, for instance through SMT solvers and the so-called hammers embedded in some proof assistants. A dual question is to know how deep one can go formalizing mathematics in a simple type theory. This work provides a first hint at an answer that should not be ruled out: simple type theory is not too simple.

@@

>>>>

In the same time, some proof assistants have made impressive progress on the algebra front. Indeed, in 2003 Laurent Chicli formalized sheaves and affine schemes in the proof assistant Coq based on a powerful dependent type theory known as the \emph{Calculus of Inductive Constructions} \cite{chiclithesis}. More recently, in 2019 Grothendieck's schemes have been formalized by the mathematician Kevin Buzzard and three students using the brave new Lean theorem prover also based on a dependent type theory\footnote{\url{https://github.com/ramonfmir/lean-scheme}} \cite{schemesinLean}. \\

@@

For our formalization we built upon Ballarin's work \cite{ballarin20} formalizing parts of Jacobson's \textit{Basic Algebra} textbook. We believe Ballarin's nice formalization could serve as a first stepping stone to a new algebra library for Isabelle. We also hope the present work could serve as a second stepping stone in that direction. Also, we seized the opportunity of formalizing schemes to partly rebuild the topology library. Our new topology library is entirely built on modern locales instead of using type classes like it is the case in \textit{HOL-Topological\_Spaces.thy}, allowing for more abstraction. In particular, our topological spaces have explicit carrier sets instead of using \textit{UNIV}. 
% snippet of code to be included here for `locale topological_space`
Within the locale for topological spaces the axiom \textit{open\_imp\_subset} enforces that an open is automatically a subset of the ambient space. This is a nice addition to the former definition, since it avoids having to add this assumption in every lemma including an open among its assumptions. Our library rebuilds basic facts about topological spaces, including for instance the definition of a topology generated by a basis, without using type classes or any deprecated feature of Isabelle. 
% snippets of code to be included here for `inductive generated_topology` and the `lemma generated_topology_is_topology` 
\\
To summarize, our work makes a triple contribution to Isabelle/HOL: a first stepping stone towards a new topology library, a second stepping stone towards a new algebra library and a formalization of schemes in algebraic geometry. In particular, our work demonstrates that intricate dependencies of algebraic structures can be managed elegantly in Isabelle. 
Actually, it seems that our experience in Isabelle to formalize schemes was much smoother than the corresponding one in Lean, since Kevin Buzzard reported on his blog the important difficulties he faced during his formalization in Lean:
\begin{quote}
	The project is completely incompatible with modern Lean and mathlib, but if you compile it then you get a sorry-free proof that an affine scheme is a scheme. During our proof we ran into a huge problem because our blueprint, the Stacks Project, assumed that $R[1/f][1/g]=R[1/fg]$, and this turns out to be unprovable for Lean’s version of equality [\dots]. The Lean community wrestled with this idea, and has ultimately come up with a beefed-up notion of equality for which the identity is now true.\footnote{\url{https://xenaproject.wordpress.com/2020/06/05/the-sphere-eversion-project/}}
\end{quote}	
For the first author of the present work, well-versed in dependent type theory, it raises the question of whether dependent types could be, at least in some situations, more of a burden than an advantage. Indeed, in the age of machine learning one may suspect there is a trade-off between the expressiveness of a type theory and one's ability to subject it to automation, for instance through SMT solvers and the so-called hammers embedded in some proof assistants. A dual question is to know how deep one can go formalizing mathematics in a simple type theory. This work provides a first hint at an answer that should not be ruled out: simple type theory is not too simple.

@@

	\bibitem{schemesinLean}
	Buzzard K., Hughes C., Lau K., Livingston A., Mir R. and Morrison S.:
	\newblock \textit{Schemes in Lean}.
	\newblock arXiv:2101.02602
